---
title: "Pesticides"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Setup
```{r echo=FALSE, include=FALSE}
# packages
require(data.table)
require(mgcv)
require(ggplot2)
require(gratia) # helper functions by Gavin Simpson
require(mgcViz) # Visualisation of GAMs

# plot theme
theme_set(theme_minimal())
```

# Data
```{r}
pest = readRDS('GAM/data/pesticides.rds')
```

# Prepare
```{r}
sit_pend = c('BW_CYY200', 'BW_CEN901', 'BW_CTA913', 'BW_CKO905',
             'SN_OBF40500', 'SN_OBF16700', 'SN_OBF32300')

pend = pest[ name == 'Pendimethalin' &
               site_id %in% sit_pend ]
```

# Add variables

```{r}
pend[ , doy := yday(date) ] # data.table function to extract day of year (doy)
pend[ , year := year(date) ]
pend[ , pa := as.numeric(as.logical(value_fin)) ] # presence-absence
```

# Task

## Model seasonal (`doy`) and long-term trends (`year`) of pendimethalin

- Model seasonal presence (`pa`) of pendimethalin
  - Does the presence change throughout the course of the year?

- Model long-term presence (`pa`) of pendimethalin
  - Does the presence change on the long run?
  
- Include sites (`site_id`) as a Random Effect in the GAM
  - Do you have to change the data type of `site_id` to use it in a model?
  - Use `bs = 're'`
  - Does this improve the model fit?
  
- Include the covariates catchment size (`area_sqkm`) & percentage agriculture (`atkis_perc`)
  - Does this improve the model fit?
  
- Predict and plot the `doy` variable of the best model


