---
title: "Pesticides"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Setup
```{r echo=FALSE, include=FALSE}
require(knitr)
require(rmarkdown)
# packages
require(data.table)
require(mgcv)
require(ggplot2)
require(gratia) # helper functions by Gavin Simpson
require(mgcViz) # Visualisation of GAMs

# plot theme
theme_set(theme_minimal())
```

# Data
```{r}
pest = readRDS('/home/scharmueller/Projects/teaching/GAM/data/pesticides.rds')
pest = readRDS('GAM/data/pesticides.rds')
```

# Prepare
```{r}
sit_pend = c('BW_CYY200', 'BW_CEN901', 'BW_CTA913', 'BW_CKO905',
             'SN_OBF40500', 'SN_OBF16700', 'SN_OBF32300')

pend = pest[ name == 'Pendimethalin' &
               site_id %in% sit_pend ]
```

# Add variables

```{r}
pend[ , doy := yday(date) ] # data.table function to extract day of year (doy)
pend[ , year := year(date) ]
pend[ , pa := as.numeric(as.logical(value_fin)) ] # presence-absence
```

# Task

## Model seasonal (`doy`) and long-term trends (`year`) of pendimethalin

- Model seasonal presence (`pa`) of pendimethalin
  - Does the presence change throughout the course of the year?

- Model long-term presence (`pa`) of pendimethalin
  - Does the presence change on the long run?
  
- Include sites (`site_id`) as a Random Effect in the GAM
  - Do you have to change the data type of `site_id` to use it in a model?
  - Use `bs = 're'`
  - Does this improve the model fit?
  
- Include the covariates catchment size (`area_sqkm`) & percentage agriculture (`atkis_perc`)
  - Does this improve the model fit?
  
- Predict and plot the `doy` variable of the best model


```{r}
# Option 1
m1 = gam(pa ~ s(doy, bs = 'cc') + s(year),
         data = pend) 
new = with(pend,
           expand.grid(doy = seq(1, 150, length.out = 100),
                       year = seq(min(year), max(year), length.out = 100)))
setDT(new) # data.table

plot(m1, pages = 1)

prd = predict(m1,
              newdata = new,
              type = 'response',
              exclude = 's(year)',
              se.fit = TRUE)
new[ , `:=`
     (fit = prd$fit,
       lwr = prd$fit - (2 * prd$se.fit),
       upr = prd$fit + (2 * prd$se.fit)) ]

ggplot(new, aes(y = fit, x = doy)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.1) +
  geom_line() +
  scale_y_continuous(limits = c(0,1))

# Option 2
pend$f1 = fitted(m1)
ggplot(pend, aes(y = f1, x = doy)) +
  geom_line()
```










