---
title: "Zooplankton"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Exercise 2

Model seasonal variation of Zooplankton densities. The data is taken from the article [Hierachical generalized additive models in ecology: an introduction with mgcv](https://peerj.com/preprints/27320/) ([GitHub-Repo](https://github.com/eric-pedersen/mixed-effect-gams)) and was originally collected by Richard Lathrop approximately bi-weekly from 1976 to 1994 ([Article](https://link.springer.com/chapter/10.1007/978-1-4612-4410-3_8).

# Setup
```{r}
require(mgcv)
require(gratia)
```


# Data

```{r}
zop = read.csv('data/zooplankton_example.csv')
head(zop)
summary(zop)
```

## Variables

day = day of the year
year
lake = name of the lake (categorical)
taxon = zooplankton taxon
density = zooplankton density (count)
density_adj = zooplankton density adjusted
density_scaled = zooplankton density scaled (to the area and volume of the net)

# Task

## Estimate seasonal variability of Cyclopoid copepods in different lakes

- use density_adj as a response
  - Which distributional family would you pick?
- use the variable day to model possible seasonl patterns
  - which smoothing basis wold you use for seasonal data?
- account for the different lakes in the model
  - as random effects
  - as factor-smooth interaction (using the `bs = 'fs'` argument)
  - as factor-smooth interaction (using the `by =` argument)
- account for inter-year variability in each lake
  - make year a factor

- Pick the best model

