---
title: "Pesticides"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Setup
```{r echo=FALSE, include=FALSE}
# packages
require(data.table)
require(mgcv)
require(ggplot2)
require(gratia)
require(mgcViz)

# plot theme
theme_set(theme_minimal())

# prj dir
prj = '/home/scharmueller/Projects/teaching/GAM'
```

# Data
```{r}
pest = readRDS(file.path(prj, 'data/pesticides.rds'))
```

# Prepare
```{r}
sit_pend = c('BW_CYY200', 'BW_CEN901', 'BW_CTA913', 'BW_CKO905',
             'SN_OBF40500', 'SN_OBF16700', 'SN_OBF32300')

pend = pest[ name == 'Pendimethalin' &
               site_id %in% sit_pend ]
```

# Add variables

```{r}
pend[ , doy := yday(date) ]
pend[ , year := year(date) ]
pend[ , pa := as.numeric(as.logical(value_fin)) ] # presence-absence
```

# Task

## Model seasonal and long-term trends of pendimethalin

- Model seasonal presence (`pa`) of pendimethalin
  - Does the presence change throughout the course of the year?

- Model long-term presence (`pa`) of pendimethalin
  - Does the presence change on the long run?
  
- Include sites (`site_id`) as a Random Effect in the GAM
  - Do you have to change the data type of `site_id` to use it in a model?
  - Does this improve the model fit?
  
- Include the covariates catchment size (`area_sqkm`) & percentage agriculture (`atkis_perc`)
  - Does this improve the model fit?
  
- Predict and plot the best model

# Models

## Model seasonality
```{r}
pe0 = gam(pa ~ s(doy, bs = 'cc'),
          data = pend,
          family = 'binomial',
          method = 'REML')
AIC(pe0)
summary(pe0)
plot(pe0)
gam.check(pe0)
```

## Model seasonality and long-term trend

```{r}
pe1 = gam(pa ~
            s(doy, bs = 'cc') +
            s(year),
          data = pend,
          family = 'binomial',
          method = 'REML')
AIC(pe1)
summary(pe1)
plot(pe1, pages = 1)
```

## Model seasonality and long-term trend including sites

```{r}
pend[ , site_id_f := as.factor(site_id) ]
pe2 = gam(pa ~
            s(doy, bs = 'cc') +
            s(year) +
            s(site_id_f, bs = 're'),
          data = pend,
          family = 'binomial',
          method = 'REML')
AIC(pe2)
summary(pe2)
plot(pe2, pages = 1)
```

## Model catchment size and agriculture therein

```{r}
k_sit = length(unique(pend$site_id_f))
pe3 = gam(pa ~
            s(doy) +
            s(year) +
            s(atkis_perc, k = k_sit) +
            s(area_sqkm, k = k_sit) +
            s(site_id_f, bs = 're'),
          data = pend,
          family = 'binomial',
          method = 'REML')
AIC(pe3)
summary(pe3)
plot(pe3, pages = 1)
```

# Predict and Plot the best model

```{r}
new = data.table(doy = seq(0, 365, length.out = 100),
                 year = 2008,
                 # atkis_perc = 0.75,
                 # area_sqkm = 100,
                 site_id_f = 'BW_CYY200')
prd = predict(pe2,
              newdata = new,
              type = 'response',
              se.fit = TRUE,
              exclude = c('s(site_id)', 's(year)'))
new[ , `:=`
    (fit = prd$fit,
      lwr = prd$fit - (1.96 * prd$se.fit),
      upr = prd$fit + (1.96 * prd$se.fit)) ]
# alternative
# ilink = family(pe2)$linkinv # function to convert from link to response scale
# new[ , `:=`
#     (fit = ilink(prd$fit),
#       lwr = ilink(prd$fit - (1.96 * prd$se.fit)),
#       upr = ilink(prd$fit + (1.96 * prd$se.fit))) ]

# plot
ggplot(new, aes(y = fit, x = doy)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.1) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = seq(1, 365, length.out = 12), labels = month.abb) +
  labs(title = 'Pendimethalin presence',
       subtitle = paste0('R2: ', round(summary(pe2)$r.sq, 2)),
       x = 'Month')
```

# Gam VIZ

```{r}
require(mgcViz)
vis.gam(pe2, # GAM object
        view = c('doy', 'year'), # variables
        plot.type = "persp", # 3D plot
        theta = 320, # horizontal rotation
        phi = 30, # phi vertical rotation
        r = 10) # zoom
```






