---
title: "Zooplankton"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Exercise 2

Model seasonal variation of Zooplankton densities. The data is taken from the article [Hierachical generalized additive models in ecology: an introduction with mgcv](https://peerj.com/preprints/27320/) ([GitHub-Repo](https://github.com/eric-pedersen/mixed-effect-gams)) and was originally collected by Richard Lathrop approximately bi-weekly from 1976 to 1994 ([Article](https://link.springer.com/chapter/10.1007/978-1-4612-4410-3_8).

# Setup
```{r}
require(mgcv)
require(gratia)
```


# Data
```{r}
zop = read.csv('GAM/data/zooplankton_example.csv')
head(zop)
summary(zop)
```

## Variables

day = day of the year
year
lake = name of the lake (categorical)
taxon = zooplankton taxon
density = zooplankton density (count)
density_adj = zooplankton density adjusted
density_scaled = zooplankton density scaled (to the area and volume of the net)

# Task

## Estimate seasonal variability of Cyclopoid copepods in different lakes

- use density_adj as a response
  - Which distributional family would you pick?
- use the variable day to model possible seasonl patterns
  - which smoothing basis wold you use for seasonal data?
- account for the different lakes in the model
  - as random effects
  - as factor-smooth interaction (using the `bs = 'fs'` argument)
  - as factor-smooth interaction (using the `by =` argument)
- account for inter-year variability in each lake
  - make year a factor

- Pick the best model

# Preparation

```{r}
zop$year_f = factor(zop$year)
zop$lake_f = factor(zop$lake)
damo = zop[ zop$taxon == 'Cyclopoid copepods', ]
```

# Model

## Only days
```{r}
zoo1 = gam(density_adj ~ s(day, bs = 'cc'),
           data = damo,
           family = Gamma(link = 'log'),
           method = 'REML')
summary(zoo1)
```

## Random effect intercept model

varying intercepts
```{r}
zoo_daG = gam(density_adj ~ s(day, bs = 'cc', k = 10) +
                            s(lake_f, bs = 're') +
                            s(lake_f, year_f, bs = 're'),
              data = damo,
              # family = gaussian(link = 'log'),
              family = Gamma(link = 'log'),
              method = 'REML')
summary(zoo_daG)
plot(zoo_daG, pages = 1)
gam.check(zoo_daG)
AIC(zoo_daG)
```

gam.check()
QQ-plot: Gamma distribution seems to fit data well, except low values
Residual vs. fitted: lowest values, absent data

## Factor smooth interaction

`bs = 'fs'` argument

varying intercepts and slopes
```{r}
zoo_daGS = gam(density_adj ~ s(day, bs = 'cc', k = 10) +
                             s(day, lake_f, k = 10, bs = 'fs') + # one basis: cont, one: RE
                             # no need to specify lake as a random effect
                             s(lake_f, year_f, bs = 're'),
               data = damo,
               family = Gamma(link = 'log'),
               method = 'REML',
               knots = list(day = c(0,365)))
summary(zoo_daGS)
plot(zoo_daGS, pages = 1)
draw(zoo_daGS)
gam.check(zoo_daGS)
AIC(zoo_daGS)
```

## Factor-smooth interaction

`by=` argument

Varying intercepts, varying Slopes, varying Wiggliness
```{r}
zoo_daGI = gam(density_adj ~ s(day, bs = 'cc') +
                             s(day, by = lake_f, bs = 'cc') +
                             s(lake_f, bs = 're') + # explicitly add lake as RE
                             s(lake_f, year_f, bs = 're'),
               data = damo,
               family = Gamma(link = 'log'),
               method = 'REML',
               knots = list(day = c(0,365)))
summary(zoo_daGI)
plot(zoo_daGI, pages = 1)
gam.check(zoo_daGI)
AIC(zoo_daGI)
```








